---
title: "Behind the Streams: What Viewers Actually Watch on Netflix"
subtitle: "Assignment 4 ETC5512"
author: Yashitta Bawa
format: html
editor: visual
message: false
warning: false
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data Details

### 🧠 Research Question

This project explores the question: **What Viewers Actually Watch on Netflix?**

The goal is to determine which types of content—such as genres, release formats, or show categories—tend to attract the highest viewership. Specifically, the analysis investigates:

-   Whether certain genres (e.g., documentaries, thrillers, reality shows) sustain more viewer attention\
-   Whether newer titles perform better than older ones\
-   What common characteristics are shared by the most-watched shows

This question was selected because streaming platforms like Netflix play a major role in everyday entertainment. Gaining insight into what makes certain content more engaging can help illuminate viewer preferences, binge-watching behavior, and content design strategies.

------------------------------------------------------------------------

### 📁 The Data

Two datasets were used in this project:

1.  **Netflix Engagement Report (Jan–Jun 2023)**\
    This dataset, published by Netflix, includes total hours watched globally for each title, along with release dates. It reflects real-world viewership patterns and provides a direct measure of audience engagement.

2.  **Netflix Titles Metadata (from Kaggle)**\
    This dataset contains descriptive information about Netflix content, including title, type (TV show or movie), duration (either runtime in minutes or number of seasons), genre categories, and release year.

These datasets were selected because they are **publicly available**, nclude the **relevant variables** needed to address the research question, and are **up-to-date** as of 2023.

They are appropriate for the intended analysis because:

-   The engagement data directly captures viewer attention

-   The metadata allows for grouping by format and genre

-   Both datasets are observational and structurally compatible after cleanings

**Licensing**:

-   The engagement report is published by Netflix and is freely downloadable from the Netflix What We Watched site

-   The titles dataset is available on Kaggle under a public license

------------------------------------------------------------------------

### 🛠️ Data Collection and Preparation

**Data Download Process:**

1.  **Netflix Engagement Report (Jan–Jun 2023):**

-   Go to: [Netflix What We Watched](https://about.netflix.com/en/news/what-we-watched-a-netflix-engagement-report)
-   Scroll down to the line **Download the report here**.
-   Click the **here** text to download the Excel file (`Netflix_Engagement_Report.xlsx`).
-   Save it into your project’s `data` folder.

2.  **Netflix Metadata (from Kaggle):**

-   Go to: [Kaggle Netflix Titles Dataset](https://www.kaggle.com/datasets/shivamb/netflix-shows)
-   Log in to Kaggle (you’ll need an account).
-   Click the blue **Download** button on the top-right corner.
-   Save the `netflix_titles.csv` file into your project’s `data` folder.

**Data Cleaning and Processing:**

-   The engagement report included a non-tabular header section, which was skipped.

-   Column names were cleaned and standardized using the `janitor` package.

-   The `title` column was standardized across both datasets by converting text to lowercase and trimming white space.

-   Only relevant variables were retained, including title, hours viewed, release date, genre, and type.

-   The two datasets were merged using the cleaned `title` column.

-   Some shows appeared multiple times in the engagement data (e.g., one row per season). These entries were retained for more granular analysis.

-   The `duration` column indicates runtime in minutes for movies and season count for TV shows, which limits direct comparisons across formats.

These steps ensured the datasets were clean, merged, and ready for structured analysis and visualization.

```{r}
# Loaded required libraries
library(tidyverse)
library(readxl)
library(here)
library(janitor)
library(scales)
library(ggridges)
library(patchwork)

# Loaded the datasets
titles <- read_csv(here("data/netflix_titles.csv")) %>% 
  clean_names()
engagement_report <- read_excel(here("data/Netflix_Engagement_Report.xlsx"), 
                                     skip = 5) %>% 
  clean_names()

# Standardised the titles for merging
titles <- titles %>%
  mutate(title_clean = str_to_lower(str_trim(title)))

engagement_report <- engagement_report %>%
  mutate(title_clean = str_to_lower(str_trim(title)),
    release_year = lubridate::year(release_date))

# Merged datasets using a left join
netflix_clean <- left_join(engagement_report, titles, by = "title_clean") %>%
  filter(!is.na(type), !is.na(duration), !is.na(listed_in)) %>%
  mutate(
    release_year = lubridate::year(release_date)) %>%
  select(title = title.x,
         release_date,
         hours_viewed,
         release_year,
         type, 
         duration, 
         listed_in)

# Added a new column with season removed
netflix_clean <- netflix_clean %>%
  mutate(base_title = str_remove(title, ": Season \\d+|: Limited Series"))

# Group by base title + genre/type and summarise
netflix_grouped <- netflix_clean %>%
  group_by(base_title, type, listed_in) %>%
  summarise(
    total_hours = sum(hours_viewed, na.rm = TRUE),
    .groups = "drop")

# Separated movie duration (in minutes) and show duration (season count)
netflix_clean <- netflix_clean %>%
  mutate(
    duration_num = parse_number(duration),
    duration_type = case_when(
      str_detect(duration, "min") ~ "minutes",
      str_detect(duration, "Season") ~ "seasons",
      TRUE ~ "unknown"))
```

------------------------------------------------------------------------

### 📊 Dataset Creation for Visualisation

**Genre Breakdown: Most Watched Categories Overall**

To identify which genres attract the highest viewership, multi-genre entries were separated, and total hours viewed were aggregated for each individual genre.

To explore trends across formats, the dataset was further grouped by both `type` (Movie or TV Show) and `genre`. The top 10 genres for each format were selected based on total viewing hours, then combined into a single dataset to allow visual comparison. While some genres appeared in both lists (e.g., Dramas, Sci-Fi & Fantasy), others were format-specific—such as Crime TV Shows or Romantic Movies.

This preparation enabled a clear side-by-side breakdown of audience preferences by format, visualised through a faceted bar chart in the blog post tab. The plot reveals which genres perform better within each format and illustrates how structural differences in content influence viewership.

```{r}
genre_summary <- netflix_clean %>%
  separate_rows(listed_in, sep = ",\\s*") %>%
  group_by(listed_in) %>%
  summarise(total_hours = sum(hours_viewed, na.rm = TRUE)) %>%
  arrange(desc(total_hours)) %>%
  slice_head(n = 10)

# Preprocess genre order
genre_summary <- genre_summary %>%
  mutate(listed_in = factor(listed_in, 
                            levels = listed_in[order(total_hours)]))
```

**Genre Preferences by Format**

To compare viewership patterns between TV shows and movies, titles were grouped by type, and summary statistics—such as mean and median hours viewed—were calculated.

To further examine genre-based viewing trends across formats, titles were grouped by both `type` and `genre`, and total viewing hours were aggregated.\

The top 10 genres for each format were then selected to showcase the most popular categories, with overlapping genres retained to allow comparison.

This data was used to create a **side-by-side faceted bar chart**, displaying the top genres by total hours viewed for **movies** on the left and **TV shows** on the right.\

The layout highlights which genres perform better within each content format.

```{r}
format_summary <- netflix_clean %>%
  group_by(type) %>%
  summarise(
    mean_hours = mean(hours_viewed, na.rm = TRUE),
    median_hours = median(hours_viewed, na.rm = TRUE),
    count = n())

# Full summary of genres by format
genre_all <- netflix_clean %>%
  separate_rows(listed_in, sep = ",\\s*") %>%
  group_by(type, genre = listed_in) %>%
  summarise(hours_viewed = sum(hours_viewed, na.rm = TRUE), .groups = "drop")

# Top 10 Movie genres
top_movie_genres <- genre_all %>%
  filter(type == "Movie") %>%
  arrange(desc(hours_viewed)) %>%
  slice_head(n = 10) %>%
  pull(genre)

# Top 10 TV genres
top_tv_genres <- genre_all %>%
  filter(type == "TV Show") %>%
  arrange(desc(hours_viewed)) %>%
  slice_head(n = 10) %>%
  pull(genre)

# Combine all 20 genres (with overlap if any)
selected_genres <- union(top_movie_genres, top_tv_genres)

# Final dataset: includes Movie + TV rows for those 20 genres
genre_by_type <- genre_all %>%
  filter(genre %in% selected_genres)
```

**Duration vs. Format**

To explore how content length varies between movies and TV shows, the `duration` column was cleaned and split into two distinct units: **minutes** (for movies) and **seasons** (for TV shows). These values were extracted numerically and stored in a new column called `duration_num`.

Entries labelled as "other" or missing type/duration values were filtered out to create a consistent dataset for comparing structural differences in content length.

The cleaned dataset was then split by format (`Movie` vs `TV Show`) and visualised using **violin plots**. Each plot displays the distribution of durations across titles, offering insight into how runtime conventions differ by format.

For instance, movies typically range from 90–120 minutes, while most TV shows cluster around 1–2 seasons. This transformation supports a focused visualisation of **duration norms** and lays the foundation for analysing their relationship with viewer engagement.

```{r}
duration_clean <- netflix_clean %>%
  mutate(
    duration_num = readr::parse_number(duration),
    duration_type = case_when(
      str_detect(duration, "min") ~ "minutes",
      str_detect(duration, "Season") ~ "seasons",
      TRUE ~ "other")) %>%
  filter(!is.na(duration_num))

# Clean duration data
duration_clean <- netflix_clean %>%
  filter(!is.na(duration), 
         !is.na(type)) %>%
  mutate(
    duration_num = case_when(
      str_detect(duration, "min") ~ parse_number(duration),
      str_detect(duration, "Season") ~ parse_number(duration),
      TRUE ~ NA_real_),
    duration_unit = case_when(
      str_detect(duration, "min") ~ "Minutes",
      str_detect(duration, "Season") ~ "Seasons",
      TRUE ~ "Other")) %>%
  filter(!is.na(duration_num), 
         duration_unit != "Other")
```
