---
title: "Behind the Streams: What Viewers Actually Watch on Netflix"
subtitle: "Assignment 4 ETC5512"
author: Yashitta Bawa
format: html
editor: visual
message: false
warning: false
---

::: panel-tabset
<!-- Task 1 Tab: Data and Documenting -->

## Data Details

### 🧠 Research Question

This project explores the question: **What Viewers Actually Watch on Netflix?**

The goal is to determine which types of content—such as genres, release formats, or show categories—tend to attract the highest viewership. Specifically, the analysis investigates:

-   Whether certain genres (e.g., documentaries, thrillers, reality shows) sustain more viewer attention\
-   Whether newer titles perform better than older ones\
-   What common characteristics are shared by the most-watched shows

This question was selected because streaming platforms like Netflix play a major role in everyday entertainment. Gaining insight into what makes certain content more engaging can help illuminate viewer preferences, binge-watching behavior, and content design strategies.

------------------------------------------------------------------------

### 📁 The Data

Two datasets were used in this project:

1.  **Netflix Engagement Report (Jan–Jun 2023)**\
    This dataset, published by Netflix, includes total hours watched globally for each title, along with release dates. It reflects real-world viewership patterns and provides a direct measure of audience engagement.

2.  **Netflix Titles Metadata (from Kaggle)**\
    This dataset contains descriptive information about Netflix content, including title, type (TV show or movie), duration (either runtime in minutes or number of seasons), genre categories, and release year.

These datasets were selected because they are **publicly available**, nclude the **relevant variables** needed to address the research question, and are **up-to-date** as of 2023.

They are appropriate for the intended analysis because:

-   The engagement data directly captures viewer attention

-   The metadata allows for grouping by format and genre

-   Both datasets are observational and structurally compatible after cleanings

**Licensing**:

-   The engagement report is published by Netflix and is freely downloadable from the Netflix What We Watched site

-   The titles dataset is available on Kaggle under a public license

------------------------------------------------------------------------

### 🛠️ Data Collection and Preparation

**Data Download Process:**

1.  **Netflix Engagement Report (Jan–Jun 2023):**

-   Go to: [Netflix What We Watched](https://about.netflix.com/en/news/what-we-watched-a-netflix-engagement-report)
-   Scroll down to the line **Download the report here**.
-   Click the **here** text to download the Excel file (`Netflix_Engagement_Report.xlsx`).
-   Save it into your project’s `data` folder.

2.  **Netflix Metadata (from Kaggle):**

-   Go to: [Kaggle Netflix Titles Dataset](https://www.kaggle.com/datasets/shivamb/netflix-shows)
-   Log in to Kaggle (you’ll need an account).
-   Click the blue **Download** button on the top-right corner.
-   Save the `netflix_titles.csv` file into your project’s `data` folder.

**Data Cleaning and Processing:**

-   The engagement report included a non-tabular header section, which was skipped.

-   Column names were cleaned and standardized using the `janitor` package.

-   The `title` column was standardized across both datasets by converting text to lowercase and trimming white space.

-   Only relevant variables were retained, including title, hours viewed, release date, genre, and type.

-   The two datasets were merged using the cleaned `title` column.

-   Some shows appeared multiple times in the engagement data (e.g., one row per season). These entries were retained for more granular analysis.

-   The `duration` column indicates runtime in minutes for movies and season count for TV shows, which limits direct comparisons across formats.

These steps ensured the datasets were clean, merged, and ready for structured analysis and visualization.

```{r}
# Loaded required libraries
library(tidyverse)
library(readxl)
library(here)
library(janitor)
library(scales)
library(ggridges)
library(patchwork)

# Loaded the datasets
titles <- read_csv(here("data/netflix_titles.csv")) %>% 
  clean_names()
engagement_report <- read_excel(here("data/Netflix_Engagement_Report.xlsx"), 
                                     skip = 5) %>% 
  clean_names()

# Standardised the titles for merging
titles <- titles %>%
  mutate(title_clean = str_to_lower(str_trim(title)))

engagement_report <- engagement_report %>%
  mutate(title_clean = str_to_lower(str_trim(title)),
    release_year = lubridate::year(release_date))

# Merged datasets using a left join
netflix_clean <- left_join(engagement_report, titles, by = "title_clean") %>%
  filter(!is.na(type), !is.na(duration), !is.na(listed_in)) %>%
  mutate(
    release_year = lubridate::year(release_date)) %>%
  select(title = title.x,
         release_date,
         hours_viewed,
         release_year,
         type, 
         duration, 
         listed_in)

# Added a new column with season removed
netflix_clean <- netflix_clean %>%
  mutate(base_title = str_remove(title, ": Season \\d+|: Limited Series"))

# Group by base title + genre/type and summarise
netflix_grouped <- netflix_clean %>%
  group_by(base_title, type, listed_in) %>%
  summarise(
    total_hours = sum(hours_viewed, na.rm = TRUE),
    .groups = "drop")

# Separated movie duration (in minutes) and show duration (season count)
netflix_clean <- netflix_clean %>%
  mutate(
    duration_num = parse_number(duration),
    duration_type = case_when(
      str_detect(duration, "min") ~ "minutes",
      str_detect(duration, "Season") ~ "seasons",
      TRUE ~ "unknown"))
```

------------------------------------------------------------------------

### 📊 Dataset Creation for Visualisation

**Genre Breakdown: Most Watched Categories Overall**

To identify which genres attract the highest viewership, multi-genre entries were separated, and total hours viewed were aggregated for each individual genre.

To explore trends across formats, the dataset was further grouped by both `type` (Movie or TV Show) and `genre`. The top 10 genres for each format were selected based on total viewing hours, then combined into a single dataset to allow visual comparison. While some genres appeared in both lists (e.g., Dramas, Sci-Fi & Fantasy), others were format-specific—such as Crime TV Shows or Romantic Movies.

This preparation enabled a clear side-by-side breakdown of audience preferences by format, visualised through a faceted bar chart in the blog post tab. The plot reveals which genres perform better within each format and illustrates how structural differences in content influence viewership.

```{r}
genre_summary <- netflix_clean %>%
  separate_rows(listed_in, sep = ",\\s*") %>%
  group_by(listed_in) %>%
  summarise(total_hours = sum(hours_viewed, na.rm = TRUE)) %>%
  arrange(desc(total_hours)) %>%
  slice_head(n = 10)

# Preprocess genre order
genre_summary <- genre_summary %>%
  mutate(listed_in = factor(listed_in, 
                            levels = listed_in[order(total_hours)]))
```

**Genre Preferences by Format**

To compare viewership patterns between TV shows and movies, titles were grouped by type, and summary statistics—such as mean and median hours viewed—were calculated.

To further examine genre-based viewing trends across formats, titles were grouped by both `type` and `genre`, and total viewing hours were aggregated.\

The top 10 genres for each format were then selected to showcase the most popular categories, with overlapping genres retained to allow comparison.

This data was used to create a **side-by-side faceted bar chart**, displaying the top genres by total hours viewed for **movies** on the left and **TV shows** on the right.\

The layout highlights which genres perform better within each content format.

```{r}
format_summary <- netflix_clean %>%
  group_by(type) %>%
  summarise(
    mean_hours = mean(hours_viewed, na.rm = TRUE),
    median_hours = median(hours_viewed, na.rm = TRUE),
    count = n())

# Full summary of genres by format
genre_all <- netflix_clean %>%
  separate_rows(listed_in, sep = ",\\s*") %>%
  group_by(type, genre = listed_in) %>%
  summarise(hours_viewed = sum(hours_viewed, na.rm = TRUE), .groups = "drop")

# Top 10 Movie genres
top_movie_genres <- genre_all %>%
  filter(type == "Movie") %>%
  arrange(desc(hours_viewed)) %>%
  slice_head(n = 10) %>%
  pull(genre)

# Top 10 TV genres
top_tv_genres <- genre_all %>%
  filter(type == "TV Show") %>%
  arrange(desc(hours_viewed)) %>%
  slice_head(n = 10) %>%
  pull(genre)

# Combine all 20 genres (with overlap if any)
selected_genres <- union(top_movie_genres, top_tv_genres)

# Final dataset: includes Movie + TV rows for those 20 genres
genre_by_type <- genre_all %>%
  filter(genre %in% selected_genres)
```

**Duration vs. Format**

To explore how content length varies between movies and TV shows, the `duration` column was cleaned and split into two distinct units: **minutes** (for movies) and **seasons** (for TV shows). These values were extracted numerically and stored in a new column called `duration_num`.

Entries labelled as "other" or missing type/duration values were filtered out to create a consistent dataset for comparing structural differences in content length.

The cleaned dataset was then split by format (`Movie` vs `TV Show`) and visualised using **violin plots**. Each plot displays the distribution of durations across titles, offering insight into how runtime conventions differ by format.

For instance, movies typically range from 90–120 minutes, while most TV shows cluster around 1–2 seasons. This transformation supports a focused visualisation of **duration norms** and lays the foundation for analysing their relationship with viewer engagement.

```{r}
duration_clean <- netflix_clean %>%
  mutate(
    duration_num = readr::parse_number(duration),
    duration_type = case_when(
      str_detect(duration, "min") ~ "minutes",
      str_detect(duration, "Season") ~ "seasons",
      TRUE ~ "other")) %>%
  filter(!is.na(duration_num))

# Clean duration data
duration_clean <- netflix_clean %>%
  filter(!is.na(duration), 
         !is.na(type)) %>%
  mutate(
    duration_num = case_when(
      str_detect(duration, "min") ~ parse_number(duration),
      str_detect(duration, "Season") ~ parse_number(duration),
      TRUE ~ NA_real_),
    duration_unit = case_when(
      str_detect(duration, "min") ~ "Minutes",
      str_detect(duration, "Season") ~ "Seasons",
      TRUE ~ "Other")) %>%
  filter(!is.na(duration_num), 
         duration_unit != "Other")
```


<!-- Task 2 Tab: Writing your blogpost -->

## Blog post

### 🎬 Motivation

With Netflix offering everything from gritty thrillers to romantic comedies, it’s natural to wonder: **what actually keeps people watching?**\
This project investigates how genre, content type, and show structure influence viewer engagement. By leveraging real-world viewing data, it uncovers patterns that help explain why some shows become binge-worthy while others fade into the background.

------------------------------------------------------------------------

### 📊 Data

The analysis is based on two combined datasets:

-   The **Netflix Engagement Report (Jan–Jun 2023)**, which lists total global viewing hours for each title.
-   A **Netflix title dataset** from Kaggle, providing additional details such as genre, content type (TV show or movie), and duration.

After merging and cleaning, each row in the final dataset represents a specific show or season along with its total hours viewed. This structured dataset allows for meaningful comparisons across genres, formats, and content lengths.

------------------------------------------------------------------------

### 📈 Analysis

#### **1. Top Genres by Total Viewing Time**

**Most-Watched Titles on Netflix (Jan–Jun 2023)**

To ground the genre and format patterns in real examples, this table showcases the five most-watched Netflix releases from the first half of 2023. It includes both movies and TV shows across a variety of genres. These titles demonstrate how diverse content—from action-packed films to true crime docuseries—can generate massive viewer engagement. Displaying this alongside the genre and duration breakdowns helps connect broader trends to familiar, recognisable shows.

```{r, echo = FALSE}
# Top 5 most-watched titles
top_titles <- netflix_clean %>%
  group_by(title, 
           type, 
           release_year) %>%
  summarise(total_hours = sum(hours_viewed, na.rm = TRUE), 
            .groups = "drop") %>%
  arrange(desc(total_hours)) %>%
  slice_head(n = 5)

# Display as table
top_titles %>%
  mutate(total_hours = scales::comma(total_hours)) %>%
  select(-release_year) %>% 
  knitr::kable(
    caption = "Table: Top 5 Most Watched Titles (Jan–Jun 2023)",
    col.names = c("Title", "Type", "Total Hours Viewed"))
```

\

To understand what types of content dominate viewer attention, total viewing hours were grouped by genre. Genres were extracted from the `listed_in` field and cleaned to remove duplicates.

```{r, echo = FALSE}
genre_plot <- ggplot(genre_summary, 
                     aes(x = listed_in,
                         y = total_hours, 
                         fill = listed_in)) +
  geom_bar(stat = "identity",
           width = 1, 
           color = "white",
           show.legend = FALSE) +
  coord_polar(clip = "off", 
              start = 0) +
  scale_y_continuous(labels = NULL) +
  labs(
    title = "Top 10 Most Watched Genres on Netflix by Hours Watched",
    x = "",
    y = "") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10, face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none",
    panel.grid = element_line(color = "grey", linetype = "dashed"),
    plot.margin = margin(1, 1, 1, 1, "cm"))
genre_plot
```

**What it shows:**\
Comedies and dramas led in viewer engagement during the first half of 2023, each accumulating over a billion hours of global watch time. Family movies and action titles also performed strongly, indicating that genres with broad, cross-generational appeal tend to attract more viewers. In contrast, niche categories like horror and sci-fi saw lower total viewing hours, though they still ranked among the top ten.

------------------------------------------------------------------------

#### **2. Genre Preferences by Format**

To compare how TV shows and movies perform across genres, titles were grouped by both `type` and `genre`, and total viewing hours were aggregated. This enabled a direct visual comparison of content preferences across formats.

```{r, echo=FALSE}
ggplot(genre_by_type, 
       aes(x = reorder(genre, hours_viewed), 
           y = hours_viewed, 
           fill = type)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~type, 
             scales = "free") +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(
    title = "Top 10 Most Watched Genres: Movies vs TV Shows",
    x = "Genre",
    y = "Total Hours Viewed") +
  theme_bw() +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.y = element_text(size = 9))
```

**What it shows:**\
This chart compares the most-watched genres on Netflix across movies and TV shows. For movies, **comedies** and **dramas** led in total viewership, followed by action, romance, and thrillers. In contrast, top-performing TV content leaned toward **mysteries**, **horror**, and **sci-fi/fantasy** — genres often associated with episodic suspense and extended story arcs.

The side-by-side layout clearly illustrates how **audience preferences vary by format**. While viewers tend to choose movies for emotionally driven or family-friendly stories, they often turn to TV shows for immersive, suspense-heavy narratives. This supports the idea that **format influences viewing behaviour**, with pacing and structure playing key roles in shaping genre appeal.

**Summary Table**\
While visual plots capture overall trends, this table highlights the five most-watched genres **within each format**. Action and comedy dominate the movie category, whereas crime and docuseries genres appear more frequently in top-performing TV content—reflecting the structural strengths of episodic storytelling in long-form viewing.

```{r, echo=FALSE}
# Explode genres
genre_by_type_top <- netflix_clean %>%
  separate_rows(listed_in, 
                sep = ",\\s*") %>%
  group_by(type, 
           genre = listed_in) %>%
  summarise(total_hours = sum(hours_viewed, 
                              na.rm = TRUE), 
            .groups = "drop") %>%
  arrange(type, desc(total_hours)) %>%
  group_by(type) %>%
  slice_head(n = 5)

# Format table nicely
genre_by_type_top %>%
  mutate(total_hours = scales::comma(total_hours)) %>%
  knitr::kable(
    caption = "Table: Top 5 Most Watched Genres by Format",
    col.names = c("Format", "Genre", "Total Hours Viewed"))
```

------------------------------------------------------------------------

#### **3. Is Longer Better?**

To examine the relationship between content length and viewer engagement, titles were grouped by format and visualised using separate violin plots—one for movie runtimes (in minutes) and another for TV show durations (in seasons).

```{r, echo=FALSE}
# Movie Plot
movie_plot <- duration_clean %>%
  filter(type == "Movie") %>%
  ggplot(aes(x = type, 
             y = duration_num,
             fill = type)) +
  geom_violin(alpha = 0.3, 
              color = NA) +
  labs(
    title = "Movie Duration Distribution",
    y = "Minutes",
    x = NULL) +
  scale_fill_manual(values = c("Movie" = "#ff6661")) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold",
                              hjust = 0.5),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none")

# TV Plot
tv_plot <- duration_clean %>%
  filter(type == "TV Show") %>%
  ggplot(aes(x = type, 
             y = duration_num, 
             fill = type)) +
  geom_violin(alpha = 0.3,
              color = NA) +
  labs(
    title = "TV Show Duration Distribution",
    y = "Seasons",
    x = NULL) +
  scale_fill_manual(values = c("TV Show" = "#00bfc4")) +
  scale_y_continuous(breaks = 1:6, 
                     limits = c(1, 6)) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold", 
                              hjust = 0.5),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    legend.position = "none")

# Combine with patchwork
movie_plot + tv_plot
```

**What it shows:**\
There is no strong or consistent link between duration and total viewership. Most **movies** clustered around the 90–120 minute range, with a few shorter and longer outliers—yet both ends of the spectrum showed varying levels of engagement. **TV shows** were predominantly limited to one-season formats, with minimal variation across the dataset. This suggests that **binge-worthiness is not solely determined by length**. Elements such as content quality, genre, release timing, and episodic structure are likely more influential in sustaining viewer interest.

------------------------------------------------------------------------

### 🎯 Conclusion

The analysis indicates that **viewer engagement on Netflix is shaped by a combination of genre, format, and structural elements**, rather than any single factor. While widely popular genres like **comedies** and **dramas** consistently attract high viewership, others—such as **horror** and **sci-fi**—perform more strongly within specific formats (e.g., TV shows).

The genre breakdown revealed clear distinctions between **movies and series**, with films leaning toward emotional or family-friendly themes, and shows favouring suspenseful, episodic storytelling. The duration analysis further showed that **most movies run between 90–120 minutes**, while **TV shows typically span 1–2 seasons**, though engagement does not necessarily increase with length.

For students exploring data storytelling, this project highlights how **public datasets can reveal cultural patterns** and viewing behaviour. Even simple preprocessing and visualisation choices—such as splitting mixed-duration fields or disaggregating genre labels—can uncover meaningful insights hidden in plain sight.

------------------------------------------------------------------------

### 📚 References

-   **Netflix Engagement Report**\
    Netflix. (2023, December 12). What we watched: A Netflix engagement report. Netflix. <https://about.netflix.com/en/news/what-we-watched-a-netflix-engagement-report>
-   **Kaggle Netflix Dataset**\
    Shivamb. (2021). Netflix movies and TV shows dataset. Kaggle. <https://www.kaggle.com/datasets/shivamb/netflix-shows>
-   **R**\
    R Core Team (2025). *R: A Language and Environment for Statistical Computing*. R Foundation for Statistical Computing, Vienna, Austria. <https://www.R-project.org/>.
-   **tidyverse**\
    Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R, Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E, Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi K, Vaughan D, Wilke C, Woo K, Yutani H (2019). “Welcome to the tidyverse.” *Journal of Open Source Software*, *4*(43), 1686. doi:10.21105/joss.01686 <https://doi.org/10.21105/joss.01686>.
-   **readxl**\
    Wickham H, Bryan J (2025). *readxl: Read Excel Files*. R package version 1.4.4, <https://CRAN.R-project.org/package=readxl>.
-   **here**\
    Müller K (2020). *here: A Simpler Way to Find Your Files*. R package version 1.0.1, <https://CRAN.R-project.org/package=here>.
-   **janitor**\
    Firke S (2024). *janitor: Simple Tools for Examining and Cleaning Dirty Data*. R package version 2.2.1, <https://CRAN.R-project.org/package=janitor>.
-   **scales**\
    Wickham H, Pedersen T, Seidel D (2023). *scales: Scale Functions for Visualization*. R package version 1.3.0, <https://CRAN.R-project.org/package=scales>.
-   **ggridges**\
    Wilke C (2024). *ggridges: Ridgeline Plots in 'ggplot2'*. R package version 0.5.6, <https://CRAN.R-project.org/package=ggridges>.
-   **patchwork**\
    Pedersen T (2024). *patchwork: The Composer of Plots*. R package version 1.3.0, <https://CRAN.R-project.org/package=patchwork>.
